---
title: "Lab2"
author: "Qiaoyu Liang"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```


```{r}
library(opendatatoronto)
library(tidyverse)
library(stringr)
# EDA
library(skimr)
# EDA
library(visdat) 
library(janitor)
library(lubridate)
library(ggrepel)
```


```{r}
res <- list_package_resources("996cfe8d-fb35-40ce-b569-698d51fc683b")
res <- res |> mutate(year = str_extract(name, "202.?"))
delay_2022_ids <- res |> filter(year==2022) |> select(id) |> pull()
delay_2022 <- get_resource(delay_2022_ids)
# make the column names nicer to work with
delay_2022 <- clean_names(delay_2022)
delay_2022 <- delay_2022 |> filter(line %in% c("BD", "YU", "SHP", "SRT"))
```



# Lab Exercises


1. Using the `delay_2022` data, plot the five stations with the highest mean delays. Facet the graph by `line`

```{r}
delay_2022 |>
  group_by(line, station) |>
  summarise(mean_delay = mean(min_delay)) |>
  arrange(-mean_delay) |>
  slice(1:5) |>
  ggplot(aes(x = station,
             y = mean_delay)) +
  geom_col() + 
  facet_wrap(vars(line), 
             scales = "free_y",
             nrow = 4) +
  coord_flip()
```



2. Using the `opendatatoronto` package, download the data on mayoral campaign contributions for 2014. Hints:
    + find the ID code you need for the package you need by searching for 'campaign' in the `all_data` tibble above
    + you will then need to `list_package_resources` to get ID for the data file
    + note: the 2014 file you will get from `get_resource` has a bunch of different campaign contributions, so just keep the data that relates to the Mayor election
  
```{r}
all_data <- list_packages(limit = 500)
list_package_resources("f6651a40-2f52-46fc-9e04-b760c16edd5c")
camps <- get_resource("5b230e92-0a22-4a15-9572-0b19cc222985")
Mayor2014 <- data.frame(camps[2])
```


3. Clean up the data format (fixing the parsing issue and standardizing the column names using `janitor`)

```{r}
Mayor2014 <- Mayor2014 %>% 
  janitor::row_to_names(1) %>%
  janitor::clean_names() 
```


4. Summarize the variables in the dataset. Are there missing values, and if so, should we be worried about them? Is every variable in the format it should be? If not, create new variable(s) that are in the right format.

```{r}
skim(Mayor2014)
```

```{r}
Mayor2014 <- Mayor2014 %>%
  mutate(contribution_amount = as.numeric(contribution_amount))
```

Yes, there are missing values in the dataset. Specifically, there are missing values for variables contributors_address, goods_or_service_desc, relationship_to_candidate, president_business_manager, authorized_representative and ward. Based on the purpose of this study, we can still do the analysis where variables with missing values are excluded. Thus, we should not be worried about the missing values in this case. Notice not every variable is in the format it should be. We notice contribution_amount is originally in character format so we change it in numeric format.

5. Visually explore the distribution of values of the contributions. What contributions are notable outliers? Do they share a similar characteristic(s)? It may be useful to plot the distribution of contributions without these outliers to get a better sense of the majority of the data. 

```{r}
Mayor2014 %>% 
  ggplot(aes(x = contribution_amount, y = ..density..)) +
  geom_histogram()
```

```{r}
Mayor2014 %>% 
  ggplot(aes(x = contribution_amount, y = ..density..)) +
  geom_histogram() + 
  scale_x_log10() + 
  labs(x= "log(contribution_amount)")
```

```{r}
Mayor2014 %>% filter(contribution_amount >= 10000)
```

```{r}
Mayor2014 %>% 
  filter(contribution_amount <= 10000) %>% 
  ggplot(aes(x = contribution_amount, y = ..density..)) + 
  geom_histogram() +scale_x_log10() +  
  labs(title = "Distribution of contributions without notable outliers", 
       x= "log(contribution_amount)")
```

Contributions that exceed 10000 can be considered as potential notable outliers. The similar characteristics are that those contributions are contributed by candidates themselves and most of them come from the Ford family.

6. List the top five candidates in each of these categories:
    + total contributions
    + mean contribution
    + number of contributions
    

```{r}
# top five candidates in total contributions
Mayor2014 %>% 
  group_by(candidate) %>% 
  summarise(total_contribution = sum(contribution_amount)) %>% 
  arrange(-total_contribution) %>% 
  slice(1:5)

# top five candidates in mean contributions
Mayor2014 %>% 
  group_by(candidate) %>% 
  summarise(mean_contribution = mean(contribution_amount)) %>% 
  arrange(-mean_contribution) %>%
  slice(1:5)

# top five candidates in number of contributions
Mayor2014 %>% 
  group_by(candidate) %>% 
  tally() %>% 
  arrange(-n) %>%
  slice(1:5)
```


7. Repeat 6 but without contributions from the candidates themselves.

```{r}
Mayor2014_7 <- Mayor2014 %>% 
  filter(contributors_name!=candidate)

# top five candidates in total contributions
Mayor2014_7 %>% 
  group_by(candidate) %>% 
  summarise(total_contribution = sum(contribution_amount)) %>% 
  arrange(-total_contribution) %>% 
  slice(1:5)

# top five candidates in mean contributions
Mayor2014_7 %>% 
  group_by(candidate) %>% 
  summarise(mean_contribution = mean(contribution_amount)) %>% 
  arrange(-mean_contribution) %>% 
  slice(1:5)


# top five candidates in number of contributions
Mayor2014_7 %>% 
  group_by(candidate) %>% 
  tally() %>% 
  arrange(-n) %>% 
  slice(1:5)
```


8. How many contributors gave money to more than one candidate?

```{r}
Mayor2014 %>% 
  group_by(contributors_name) %>% 
  distinct(candidate) %>%
  tally() %>% 
  filter(n > 1) %>% 
  nrow()
```

184 contributors gave money to more than one candidate.
